---
title: "NYPD Project"
author: "Desiree Disco"
date: "2025-04-25"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```
## Project Overview

I analyzed the NYPD Shooting Incident Data for NYC and looked at the characteristics of the perpetrators of murder victims and the characteristics of murder victims. I also looked at which boroughs have the most shootings and which months out of the year have the most shootings. I then applied a linear model to determine if there was an overall increase or decrease in shootings and murders. 


## Data Description

For my analysis, I used the NYPD Shooting Incident Data (Historic) which is a list of shooting incidents that occurred in NYC going back from 2006 through the end of the previous calendar year. The data was obtained from https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv.


## Research Questions

Are there some years that have murders with more unknown perpetrators than identified perpetrators?

What is the age, sex, and race characteristics of the perpetrators of murder victims?

What is the age, sex, and race characteristics of the murder victims?

What months of the year have the highest number of shootings?

Which boroughs experience the highest number of shootings?

Is the number of shootings/fatalities projected to increase or decrease?


## Environment Setup
```{r import_packages}
library(tidyverse)
```

## Read Data

Download the latest version of the dataset and read data into a data frame.
```{r get_data}
# load data into data frame
url_in = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
data <- read_csv(url_in)
```

## Inspect Data

Before performing the analysis, I looked at the data types in the dataset and viewed a portion of the dataset.

```{r inspect_data}
# inspect data
glimpse(data)
```

## Clean Data

I checked for duplicate values and cleaned up the perpetrator and victim columns for missing values. There are duplicated INCIDENT_KEY values; however, according to the data footnotes, the duplicated INCIDENT_KEY values represent incidents involving multiple victims and do not indicate duplicate entries.

```{r clean_data}
# check for duplicates
data_distinct <- data %>%
  distinct(INCIDENT_KEY, .keep_all = TRUE)

# count number of duplicate INCIDENT_KEY entries
num_duplicates <- nrow(data) - nrow(data_distinct)
cat("Number of duplicate INCIDENT_KEY entries:", num_duplicates,"\n")

# null and blank values with 'UNKNOWN'
data$PERP_AGE_GROUP <- replace(data$PERP_AGE_GROUP, data$PERP_AGE_GROUP == '(null)', 'UNKNOWN')
data$PERP_SEX <- replace(data$PERP_SEX, data$PERP_SEX == '(null)', 'UNKNOWN')
data$PERP_RACE <- replace(data$PERP_RACE, data$PERP_RACE == '(null)', 'UNKNOWN') 
data <- data %>% replace_na(list(PERP_AGE_GROUP = 'UNKNOWN', PERP_SEX = 'UNKNOWN', PERP_RACE = 'UNKNOWN'))

# replace M, F, U to male, female, unknown in perp_sex and vic_sex
data$PERP_SEX <- replace(data$PERP_SEX, data$PERP_SEX == 'M', 'MALE')
data$PERP_SEX <- replace(data$PERP_SEX, data$PERP_SEX == 'F', 'FEMALE')
data$PERP_SEX <- replace(data$PERP_SEX, data$PERP_SEX == 'U', 'UNKNOWN')
data$VIC_SEX <- replace(data$VIC_SEX, data$VIC_SEX == 'M', 'MALE')
data$VIC_SEX <- replace(data$VIC_SEX, data$VIC_SEX == 'F', 'FEMALE')
data$VIC_SEX <- replace(data$VIC_SEX, data$VIC_SEX == 'U', 'UNKNOWN')


# define mistakes in perp_age_group
mistakes_PERP <- c('1020', '224', '940', '1028', '2021')
mistakes_VIC <- c('1022')
 
# replace mistakes with UNKNOWN
data$PERP_AGE_GROUP <- replace(data$PERP_AGE_GROUP, data$PERP_AGE_GROUP %in% mistakes_PERP, 'UNKNOWN')
data$VIC_AGE_GROUP <- replace(data$VIC_AGE_GROUP, data$VIC_AGE_GROUP %in% mistakes_VIC, 'UNKNOWN')

```

## Tidy Data

I added some columns to the data relating to the date and if the perpetrator was identified. I dropped the columns referring to coordinates.

```{r tidy_data}
# drop columns for coordinates and incident key
data <- data %>%
  select(-c(INCIDENT_KEY,X_COORD_CD,Y_COORD_CD,Latitude,Longitude,Lon_Lat))

# add columns for year and month and perpetrator identified
data <- data %>%
  mutate(date = mdy(OCCUR_DATE)) %>%
  mutate(year = year(date), month = month(date)) %>%
  mutate(Unknown_perp = case_when(PERP_AGE_GROUP == 'UNKNOWN' & PERP_RACE == 'UNKNOWN' & PERP_SEX == 'UNKNOWN' ~ "Unknown Perpetrator", TRUE ~ 'Perpetrator Identified')) %>%
  select(BORO, date, year, month, STATISTICAL_MURDER_FLAG, PERP_AGE_GROUP, PERP_SEX, PERP_RACE, Unknown_perp, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)

# get the years covered in the dataset to be used in future graphs
min_year = min(data$year)
max_year = max(data$year)
year_lst = c(min_year:max_year)
```

## Perpetrator Analysis

First, I looked at the percentage of perpetrators unknown verse identified of shootings that are classified as murders. Then I looked at the demographic characteristics of the perpetrators of shootings classified as murders such as age, sex, and race.

```{r perp_analysis}

# grouping data by identified or unknown and totaling the number of murders
grouped_data <- data %>%
  group_by(Unknown_perp, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# group to get the year total
group_murders_year_total <- grouped_data %>%
  group_by(year) %>%
  summarize(total_year_murders = sum(murders))

# merge the tables by year
group_murders_perp_id = merge(x = grouped_data, y = group_murders_year_total, by = 'year', all.x = TRUE)
group_murders_perp_id <- group_murders_perp_id %>%
  mutate(percent = paste0(round(murders/total_year_murders * 100), '%'))

ggplot(data = group_murders_perp_id, aes(x = year, y = murders, fill = Unknown_perp, label = percent)) +
  geom_bar(stat = "identity") +
  geom_text(size = 3, position = position_stack(vjust = .5), angle = 90) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) +
  labs(title = 'Percentage of Murders by Unknown/Identified Perpetrators', x = 'Year',
       y = 'Number of Murders', fill = 'Unknown/Identified')

# grouping data by age range and totaling the number of murders
group_murders_year_PERP_AGE <- data %>%
  group_by(PERP_AGE_GROUP, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# line graph with line type and color by age range
ggplot(data = group_murders_year_PERP_AGE, mapping = aes(x = year, y = murders)) + 
  geom_line(aes(group=PERP_AGE_GROUP, color=PERP_AGE_GROUP, linetype=PERP_AGE_GROUP)) +
  geom_point(aes(color=PERP_AGE_GROUP)) +
  scale_linetype_manual(values=c('solid', 'solid', 'solid', 'solid', 'solid', 'dotted')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) + 
  labs(title = 'Annual Murder Counts by Perpetrator\'s Age', x = 'Year', 
       y = 'Number of Murders', color = 'Age Group', linetype = 'Age Group') + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

# grouping data by sex and totaling the number of murders
group_murders_year_PERP_SEX <- data %>%
  group_by(PERP_SEX, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# line graph with line type and color by sex
ggplot(data = group_murders_year_PERP_SEX, mapping = aes(x = year, y = murders)) + 
  geom_line(aes(group=PERP_SEX, color=PERP_SEX, linetype=PERP_SEX)) +
  geom_point(aes(color=PERP_SEX)) +
  scale_linetype_manual(values=c('solid', 'solid', 'dotted')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) + 
  labs(title = 'Annual Murder Counts by Perpetrator\'s Sex', x = 'Year', 
       y = 'Number of Murders', color = 'Sex', linetype = 'Sex') + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

# grouping data by race and totaling the number of murders
group_murders_year_PERP_RACE <- data %>%
  group_by(PERP_RACE, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# line graph with line type and color by race
ggplot(data = group_murders_year_PERP_RACE, mapping = aes(x = year, y = murders)) + 
  geom_line(aes(group=PERP_RACE, color=PERP_RACE, linetype=PERP_RACE)) +
  geom_point(aes(color=PERP_RACE)) +
  scale_linetype_manual(values=c('solid', 'solid', 'solid', 'solid', 'dotted', 'solid', 'solid')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) + 
  labs(title = 'Annual Murder Counts by Perpetrator\'s Race', x = 'Year', 
       y = 'Number of Murders', color = 'Race', linetype = 'Race') + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

```

## Victim Analysis

Next, I looked at the demographic characteristics of victims of shootings classified as murders such as age, sex, and race.

```{r vic_analysis}

# grouping data by age range and totaling the number of murders
group_murders_year_VIC_AGE <- data %>%
  group_by(VIC_AGE_GROUP, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# line graph with linetype and color by age range
ggplot(data = group_murders_year_VIC_AGE, mapping = aes(x = year, y = murders)) + 
  geom_line(aes(group = VIC_AGE_GROUP, color = VIC_AGE_GROUP, linetype = VIC_AGE_GROUP)) +
  geom_point(aes(color = VIC_AGE_GROUP)) +
  scale_linetype_manual(values = c('solid', 'solid', 'solid', 'solid', 'solid', 'dotted')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels = year_lst, breaks = year_lst) + 
  labs(title = 'Annual Murder Counts by Victim\'s Age', x = 'Year', 
       y = 'Number of Murders', color = 'Age Group', linetype = 'Age Group') + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

# grouping data by sex and totaling the number of murders
group_murders_year_VIC_SEX <- data %>%
  group_by(VIC_SEX, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# line graph with linetype and color by sex
ggplot(data = group_murders_year_VIC_SEX, mapping = aes(x = year, y = murders)) + 
  geom_line(aes(group = VIC_SEX, color = VIC_SEX, linetype = VIC_SEX)) +
  geom_point(aes(color = VIC_SEX)) +
  scale_linetype_manual(values = c('solid', 'solid', 'dotted')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels = year_lst, breaks = year_lst) + 
  labs(title = 'Annual Murder Counts by Victim\'s Sex', x = 'Year', 
       y = 'Number of Murders', color = 'Sex', linetype = 'Sex') + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

# grouping data by race and totaling the number of murders
group_murders_year_VIC_RACE <- data %>%
  group_by(VIC_RACE, year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# line graph with linetype and color by race
ggplot(data = group_murders_year_VIC_RACE, mapping = aes(x = year, y = murders)) + 
  geom_line(aes(group=VIC_RACE, color=VIC_RACE, linetype=VIC_RACE)) +
  geom_point(aes(color=VIC_RACE)) +
  scale_linetype_manual(values=c('solid', 'solid', 'solid', 'solid', 'dotted', 'solid', 'solid')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) + 
  labs(title = 'Annual Murder Counts by Victim\'s Race', x = 'Year', 
       y = 'Number of Murders', color = 'Race', linetype = 'Race') + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

```

## Overall Trend

Then I looked at the number of shootings by borough and looked at the seasonality of shootings. Lastly, I applied linear regression models to shootings and murders by year to look at the overall trend and see if either model was statistically significant.

```{r trends}
# group by month to look at seasonality of shootings
group_total_month <- data %>%
  group_by(month) %>%
  summarise(shootings = n(), .groups = "drop_last") %>%
  mutate(avg_shootings_per_month = shootings / n_distinct(data$year))

# plot the seasonality trends for shootings
ggplot(data = group_total_month, mapping = aes(x = month, y = avg_shootings_per_month)) + 
  geom_bar(stat = "identity", fill = 'steelblue') +
  labs(title = 'Average Shootings per Month for all NYC Boroughs', x = 'Month', y = 'Average shootings per month') +
  scale_x_discrete(limits = c('1','2','3','4','5','6','7','8','9','10','11','12'), labels = c(
      '1' = "Jan",
      '2' = "Feb",
      '3' = "Mar",
      '4' = "Apr",
      '5' = "May",
      '6' = "Jun",
      '7' = "Jul",
      '8' = "Aug",
      '9' = "Sep",
      '10' = "Oct",
      '11' = "Nov",
      '12' = "Dec"
    )) + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

# group by borough and year to look at which borough has the most shootings per year
group_by_boro_year <- data %>%
  group_by(BORO, year) %>%
  summarise(shootings = n(), .groups = "drop_last")

# plot the breakout of shootings by borough
ggplot(data = group_by_boro_year, mapping = aes(x = year, y = shootings)) + 
  geom_line(aes(group=BORO, color=BORO)) +
  geom_point(aes(color=BORO)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) + 
  labs(title = 'Number of Shootings by Borough per Year', x = 'Year', 
       y = 'Number of Shootings', color = 'Boroughs') + 
  theme(plot.title = element_text(hjust = 0.5, vjust = 1))

# group shootings per year
group_shootings_year <- data %>%
  group_by(year) %>%
  summarise(shootings = n(), .groups = "drop_last")

# group murders per year
group_murders_year <- data %>%
  group_by(year) %>%
  summarise(murders = sum(STATISTICAL_MURDER_FLAG == 'TRUE'), .groups = "drop_last")

# linear regression of shootings per year
mod_shootings <- lm(shootings ~ year, data = group_shootings_year)
summary(mod_shootings)
shootings_by_year_w_pred <- group_shootings_year %>% mutate(pred_shooting = predict(mod_shootings))

# linear regression of murders per year
mod_murder <- lm(murders ~ year, data = group_murders_year)
summary(mod_murder)
murders_by_year_w_pred <- group_murders_year %>% mutate(pred_murder = predict(mod_murder))

# combine the shootings and murders linear regression models
combined_by_year_model <- shootings_by_year_w_pred %>%
  full_join(murders_by_year_w_pred)

# set legend colors for the visualization
legend_colors <- c('Actual Shootings' = 'darkred', 'Predicted Shootings' = 'darkorange', 'Actual Murders' = 'darkslateblue', 'Predicted Murders' = 'deepskyblue')

# plot both the shootings and murders linear regression models
combined_by_year_model %>%
  ggplot() +
  geom_line(aes(x=year, y=shootings, color = 'Actual Shootings')) +
  geom_line(aes(x=year, y=pred_shooting, color = 'Predicted Shootings')) +
  geom_line(aes(x=year, y=murders, color = 'Actual Murders')) +
  geom_line(aes(x=year, y=pred_murder, color = 'Predicted Murders')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_x_continuous(labels=year_lst,breaks=year_lst) + 
  labs(title = 'Actual vs. Predicted Shootings and Murders in NYC', x = 'Year', y = 'Number of Shootings/Murders', color = "Actual v. Predicted of Shootings/Murders") + 
  scale_color_manual(values = legend_colors) + theme(plot.title = element_text(hjust = 0.5, vjust = 1))

```

## Conclusion

My analysis of the NYPD Shooting Incident Data (Historic) dataset shows several trends and patterns related to gun violence in New York City.

**Victim and Perpetrator Demographics**
Black males aged 25–44 are more likely to be victims of gun violence. Similarly, perpetrators are predominantly Black males aged 18–44. 

**Seasonal Trends**
Gun violence displays a seasonal pattern, with a higher average number of shootings during the summer months. The seasonal spike may be due to an increase in outdoor activity as well as an increase in social activity during the warmer months.

**Geographic Distribution by Borough**
Staten Island has maintained a relatively consistent number of shootings year over year. In contrast, the remaining four boroughs (The Bronx, Brooklyn, Manhattan, and Queens) followed a similar pattern of increase and decline. The Bronx and Brooklyn reported the highest number of shootings, therefore, could benefit from intervention efforts.

**Temporal Trends**
The dataset shows a decrease in the number of shootings between 2016 and 2019, with a similar downward trend observed again in 2023. The reductions might have been influenced by external factors, such as public safety initiatives, community-based programs, or increased law enforcement. Further inquiry would be valuable to identify the specific drivers behind the declines and assess their long-term effectiveness in reducing gun violence.

**Predictive Modeling**
Based on the linear regression models to predict shooting and murder numbers, both outcomes are projected to decline over time. The model for shootings was found to be statistically significant which is supported by a low p-value, implying that it can reliably predict future trends. In contrast, the model for predicting murders had a higher p-value and is therefore not statistically significant and may not reliably predict future trends.

## Bias

There are several areas where potential bias in the dataset may exist.

**Data Collection Bias**
Since the data is collected by the City of New York, over time some of the data that was collected one year may not be collected another year. Human errors during data entry are a possible source of inaccuracy.

**Demographic Bias**
There is a large amount of missing data for perpetrator demographic information such as age, sex, and race. The lack of data may introduce demographic bias and limit the accuracy of any conclusions regarding perpetrators.

**Statistical Murder Flag**
Based on the dataset footnotes STATISTICAL_MURDER_FLAG are shootings that resulted in the victim’s death which would be counted as a murder. However, the dataset classification is not a legal determination of murder or a guilty verdict.
 

